---
title: "Market access model"
#author: "Sebastian Palmas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

# Market access

*Modified from eHandbook script in reagro* 

This access layer is used in the maize_price_farmgate.rmd and in the fertilizer_price.rmd scripts.

## Transportation cost surface 
### Elevation and Slope

```{r slope, message=FALSE}
library(terra)
tza_alt <- rast("../data/CGIAR-SRTM/srtm_TZA.tif")
tza_slope <- rast("../data/CGIAR-SRTM/srtm_slope_TZA.tif")
tza_slope <- 2*pi*tza_slope/100#convert from percentage to radian
plot(tza_slope, main="Slope (radians)")
```

We use the slope layer obtained above to create a decay coefficient that governs how much the slope impacts the speed and that we will apply to each grid cell in the cost surface. We use a decay coefficient of 1.5.

```{r}
decay <- 1.5
slope_cost <- exp( decay * tan(tza_slope) )
names(slope_cost) <- "slope_cost"
plot(slope_cost, main="Slope cost")
```

We will use the slope layer as template for the other rasters that we will create.


### Roads


```{r plotRoads}
roads <- reagro::reagro_data("tza_roads")
roads <- project(roads, crs=crs(tza_slope))
plot(tza_slope)
lines(roads)
lines(roads[roads$highway == "secondary", ], lwd=2, col="blue")
lines(roads[roads$highway == "primary", ], lwd=4, col="red")
```

#### Road cost surface

The below takes a long time to run. It may also fail because of the relative fraility of the OSM servers (or rather the enourmous number of requests that they get). Road data is in `reagro`.


```{r rasterize}
roadtypes <- c("primary", "secondary", "tertiary")
i <- match(roads$highway, roadtypes)
roads$speed <- c(0.001, 0.0015, 0.002)[i]  #speed in the different highways (min/m)

rd_cost <- terra::rasterize(roads, tza_slope,
                            values="speed",
                            filename="../results/tif/road_cost.tif",
                            overwrite=TRUE,
                            wopt=list(names="road_cost"))

#rd_cost <- rast("../results/tif/rdcost.tif")

#I think the rasterize function is not working properly for terra right now. I'll reclassify manually.
#first i need the reclassify matrix

road_speeds <- cbind(roads=c(1,2,3), speeds = c(0.001, 0.0015, 0.002))

rd_cost2 <- classify(x=rd_cost,
                     rcl = road_speeds,
                     filename="../results/tif/access/road_cost2.tif",
                     overwrite=TRUE)

plot(rd_cost2, main="Travel cost (min/m)")
```


### Land Cover

```{r}
tza_lc <- reagro::reagro_data("TZA_globcover")
plot(tza_lc, main = "GLOBCOVER 2009 v2.3 land classes")
```

As mentioned above, creating a travel cost surface depending on land cover requires an associated travel cost for each land class. urveys, such as the one the Weiss et al., (2018) paper can be performed to obtain estimates of how long it takes (min/m) individuals to traverse each land cover type. For now, we assign some travel cost values to the different land cover classes in Tanzania.

```{r, echo = FALSE}
library(knitr)
text_tbl <- data.frame(
  Value = c(40,50,70,160,170,190,200,210,220),
  LandClass = c("Closed to open (>15%) broadleaved evergreen or semi-deciduous forest (>5m)",
                "Closed (>40%) broadleaved deciduous forest (>5m)",
                "Closed (>40%) needleleaved evergreen forest (>5m)",
                "Closed to open (>15%) broadleaved forest regularly flooded (semi-permanently or temporarily) - Fresh or brackish water",
                "Closed (>40%) broadleaved forest or shrubland permanently flooded - Saline or brackish water",
                "Artificial surfaces and associated areas (Urban areas >50%)",
                "Bare areas",
                "Water bodies",
                "Permanent snow and ice"),
  travel_speeds=c(0.04, 0.04, 0.04, 0.03, 0.05, 0.01, 0.01, 0.11, 0.13)
)

kable(text_tbl)
```

We need to reclassify the values and assign speeds in each land class.

```{r}
rc <- data.frame(from=as.vector(unique(tza_lc)), to=0.02)
rc$to[rc$from %in% c(190,200)] <- 0.01
rc$to[rc$from == 160] <- 0.03
rc$to[rc$from %in% c(40,50,70)] <- 0.04
rc$to[rc$from == 170] <- 0.05
rc$to[rc$from == 210] <- 0.11
rc$to[rc$from == 220] <- 0.13

#reclassifying
tza_lc_cost <- classify(tza_lc, rc) 
```

We need to warp to match the other layers.

```{r}
lc_cost <- warp(tza_lc_cost, tza_slope,
                method="ngb",
                filename="../results/tif/access/lc_cost.tif",
                overwrite=TRUE,
                wopt=list(names="lc_cost"))
plot(lc_cost, main = "Travel costs (min/m) depending on land cover class")
```


### Combining all travel costs

Now that we have all travel cost surfaces what we will use, we then need to combine them into a single travel cost layer that keeps only the minimum cost associated of moving though each grid cell. In this case, we combine the three cost layers in a SpatRaster object and apply the `min` function to obtain the minimum value in each grid cell.

```{r}
# Combine the cost layers
all_cost <- c(rd_cost2, lc_cost)

#getting the minimum value of each grid cell
cost <- min(all_cost, na.rm=TRUE)*slope_cost

writeRaster(cost, filename = "../results/tif/access/cost.tif", overwrite=TRUE)

#areas near mountains have very high slope resulting in really high costs. I added a zlim value for better representation

plot(cost, main="Final cost layer (min/m)", zlim=c(0,0.2))
```


## Market access

With the cost surface, we can compute market access by calculating the accumulated least cost surface from market locations. In other words, each pixel will have an associated cost of getting to a market center.

Distances within a raster can be computed with the [gdistance](https://cran.r-project.org/web/packages/gdistance/index.html) R package using the  `accCost` function.

When using least cost path analysis, the eight neighbors of a raster pixel are evaluated and the generated path moves to the cells with the smallest accumulated or cost value. This process is repeated multiple times until the source and destination are connected

Now with the location of the cities, we can calculate the access cost using `accCost` function in the `gdistance` package.

We first need to create a transition object from the cost rasters using the `transition` function. We read the cost layer produced above usign `raster::raster` because the transition function requires a `RasterLayer` object.

The transition object represent the weights given to each connection between pixels (nodes). In `gdistance`, the transition matrix is expected to have conductance values. The conductance

```{r}
# Combine the cost layers
library(gdistance)

cost <- raster("../results/tif/access/cost.tif")  #min/m

#converting to conductance values with correct units
conductance <- 1/cost

#Creating a transition object 
trans <- transition(conductance,transitionFunction=mean, directions= 8)
```

Because the transition layer is based on a lon-lat projection and covers a large area, it requires geocorrection for map distortion as well as for diagonal connections between grid cells.
```{r}
trans <- geoCorrection(trans, type="c")

#writing the transition object to use in maize_price_farmgate.rmd analysis.
save(trans, file="../results/tif/access/transition.rda")


```

With the transition object, we can now calculate access using the `accCost` function and supplying the coordinates from which to calulate.  

In this code below, we make create a spatialpoint object with locations cities (>50000 population) in Tanzania.

```{r}
#cities <- vect("F:/Work/Tanzania/ne_10m_populated_places_TZA.shp") %>% project(crs=crs(tza_slope)) %>% as.data.frame()

cities <- vect("F:/Work/Tanzania/ne_10m_populated_places_TZA.shp") %>% project(crs=crs(tza_alt))
geom(cities)

#getting the projected coordinates and the populatoin
cities <- cbind(x=geom(cities)$x, y=geom(cities)$y, POP_MAX=cities$POP_MAX) %>% as.data.frame()

#keep only large cities
cities <- cities %>% dplyr::filter(POP_MAX>=50000)

#convert to spatial points needed in gdistance
cities <- SpatialPoints(cbind(cities$x, cities$y))

#Estimating access of each pixel to the nearest 50k city
A <- accCost(trans, fromCoords=cities) 

writeRaster(A, filename = "../results/tif/access/accCost50k.tif", overwrite=TRUE)

plot(A, zlim=c(0,4000),
     main="Access to markets (cities > 50k) in Tanzania (min)")
points(cities)

```
