---
title: "Estimate farmgate prices for Tanzania"
author: "Sebastian Palmas"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---


The price of maize in the farms is lower in areas farther away from markets because the cost of transportation of the production. In this case, we assume that for each km, the cost of moving maize is 0.01 USD/kg. Therefore, the price on farms will be the maize price at the nearest market minus the transportation cost from the farm to the market.

The market location and their output value will serve as end points for a cumulative cost analysis.

To create travel time from each market to each pixel, we will use the transition object created in the `access.rmd` script.

## Packages and scripts
```{r}
rm(list=ls())
gc()
library(gdistance)
library(raster)
library(terra)
library(magrittr)
```

##Template and projection
We will the common projection. We can use the one stored in the altitude raster.
```{r}
tza_alt <- rast("../data/CGIAR-SRTM/srtm_TZA.tif")
```


#Travel time to each market
In here, we compute travel time from each market to each pixel in the area using the transition object created in the `access.rmd` file.

```{r}
#Loading trans object that was created in the access.rmd file
load("../results/tif/access/transition.rda") 

plot(raster(trans), main='Transition matrix (pixel/min?)')
```

Towns were obtained from the GRUMP database: Center for International Earth Science Information Network - CIESIN - Columbia University, CUNY Institute for Demographic Research - CIDR, International Food Policy Research Institute - IFPRI, The World Bank, and Centro Internacional de Agricultura Tropical - CIAT. 2017. Global Rural-Urban Mapping Project, Version 1 (GRUMPv1): Settlement Points, Revision 01. Palisades, NY: NASA Socioeconomic Data and Applications Center (SEDAC). https://doi.org/10.7927/H4BC3WG1. Accessed 23 10 2019.

```{r}
grump <- rgdal::readOGR("F:/Work/GRUMP/grump-v1-settlement-points-rev01-shp/global_settlement_points_v1.01_TZA_CitiesAdded.shp") %>% 
  spTransform(CRSobj = crs(tza_alt))
```


## Getting output prices at towns centers
We first need the predicted maize prices at market locations. We also need to bring the maize_price raster to 1km resolution by aggregating using a mean of the cells (the original resolution is 250m, we need to aggregate by a factor of 4 to reach 1km).

```{r, eval=FALSE}
maize_price <- rast("F:/Work/MaizePricePredictions/TZA/pred_maize_price1_USD.tif")  # from maize_price_market.rmd

#Converting to 5x5km resolution
maize_price <- warp(maize_price, tza_alt, filename="../data/prices/pred_maize_price1_USD_5km.tif", overwrite=TRUE)
```

```{r}
maize_price <- raster("../data/prices/pred_maize_price1_USD_5km.tif") %>% projectRaster(crs=crs(tza_alt))
plot(maize_price, main="Market maize prices in Tanzania")
points(grump)

```


```{r}
maize_price_grump <- extract(maize_price, grump) 

#the extract function is returning some NA values, I will complete the list with a mean of the other values
maize_price_grump[is.na(maize_price_grump)] <- mean(maize_price_grump, na.rm=TRUE)

#We'll export the predicted price for the markets for validation
market_price <- cbind(grump$Name1, round(maize_price_grump, digits = 4))
write.csv(market_price, file = "../data/prices/maize_market_price.csv")
```

## Calculting farmgate prices
We then calculate the cost of access of each pixel to the each market. This cost of access will have units of 
1) Calculating time to market
2) Use the formula of: marketprice(USD/kg) - 0.01(USD/kg/hr) * time to market (hr) to calculate the prices (USD/kg) in each pixel.
```{r}
for(point_n in seq_along(grump)){
  #point_n <- 60 # to test
  if(!is.na(raster::extract(raster(trans), grump[point_n,]))){  #there are some points that do not lie within the extent of the data. No idea why
    #time to market in hours
    time_to_market <- accCost(trans, fromCoords=grump[point_n,])/60  
    #plot(time_to_market)
    #transportation cost
    tr_cost <- 0.01 #USD/hr
    
    farmgate_price <- maize_price_grump[point_n] - time_to_market*tr_cost
    #plot(farmgate_price)
    
    writeRaster(farmgate_price,
                filename=paste0("../data/prices/maize_farmgate_price/maize_farmgate_price_point_n/maize_farmgate_price",point_n,".tif"),
                overwrite=TRUE)
  }
  #acc[is.infinite(acc)] <- NA
  #to convert from m to km
}

plot(time_to_market, main="Time to market #77 (hr)")
plot(farmgate_price, main="Farmgate price to market #77 (USD/kg)")
```


## What is the maximum cost between all market prices
Farmers will sell their produce to the highest bidder. Picking the maximum price across the possibilities. They won't send their produce very far away and lose money.
```{r}
#list all raster files
files <- list.files(path = "../data/prices/maize_farmgate_price/maize_farmgate_price_point_n/",
                    pattern = ".tif", full.names=TRUE)

#combine all rasters
farmgate_prices <- rast(files)

#getting the maximum price
farmgate_price <- max(farmgate_prices, na.rm=TRUE)
writeRaster(farmgate_price, filename="../data/prices/maize_farmgate_price/maize_farmgate_price.tif", overwrite=TRUE)

plot(farmgate_price, main ="Maximum available farm price (USD/ha)")


```
