---
title: "Results of ex-ante simulation"
author: "Sebastian Palmas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = getwd())
```

# Packages
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(directlabels)
library(ggpubr)   # for theme and ggarrange package
library(grid)  #for textGrob
library(gridExtra)  #for grid.arrange
library(terra)
library(RColorBrewer) 
library(rgdal)
library(tidyverse)
```


Other needed scripts:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
source('../code/fitness.R')

#Colorblind safe palette
cbbPalette <- c("#FFFFFF", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

```

# Framework
We developed a spatial framework that integrates biophysical and socio-economic variables over large spatial scales useful for ex-ante impact assessment of technologies. The framework data requirements depend on the models used within it. In this framework, soil nutrient and maximum attainable maize yields maps enter a crop model parameterized to predict maize yield under different fertilization scenarios.

## Yield model
The yield prediction model 

## Prices
Using this predicted yield, we calculated the profitability of multiple fertilization scenarios from the crop sale of the predicted yields using the maize prices predictions and substracting the fertilizer price from the scenario fertilizer amount. The simulation uses fertilizer and maize price maps to estimate total fertilization cost and revenue to then estimate the profitability of the maize production.

### Input prices
For this analysis, we used a price for Nitrogen of 1 USD/kg.

### Output prices
Using maize prices collected at known markets across Tanzania in 2015, we predicted market prices across Tanzania using predictors such as location, weather elements, soil properties and other variables capturing aspects of market access and population distribution (e.g. distance to main roads, population density. All variables are described in Table A1. We modeled the preduction using a random forest algorithm using R 3.5.1 from the `randomForest` package.


The model for maize market price had high fitness with the maize price from known locations (Figure X).
```{r, echo=FALSE, warning=FALSE, fig.height=3, fig.width=3,, fig.cap="Figure X. Observed vs predicted maize market prices from a Random Forest model."}
maize_farmgate_price <- rast("../data/prices/maize_farmgate_price/maize_farmgate_price.tif") #to convert from Tsh/kg to USD/kg 2013 prices

if(.Platform$OS.type == "unix") {
  maize_price_table <- read.csv("/media/sp/SPHD/Work/LSMS/TZA/maize_prices_2015/local_maize_prices_2015.csv")
} else {
  maize_price_table <- read.csv("F:/Work/LSMS/TZA/maize_prices_2015/local_maize_prices_2015.csv")
}

maize_price_vect <- vect( cbind(maize_price_table$lon_modified, maize_price_table$lat_modified), atts=maize_price_table,
                          crs="+proj=longlat +datum=WGS84 +no_defs")
maize_price_vect <- terra::project(maize_price_vect, crs=crs(maize_farmgate_price))

maize_price_table$maize_price_pred <- terra::extract(maize_farmgate_price, maize_price_vect, drop=TRUE)

maize_price_table$maipkgUSD_kg <- maize_price_table$maipkg / 1598 #to convert from TZS to USD

rsq1 <- rsq(obs = maize_price_table$maipkgUSD_kg, pred=maize_price_table$maize_price_pred) %>% round(digits=2)
RMSE1 <- RMSE(obs = maize_price_table$maipkgUSD_kg, pred=maize_price_table$maize_price_pred) %>% round(digits=2)

p <- ggplot(data=maize_price_table, aes(x=maipkgUSD_kg, y=maize_price_pred))+
  geom_abline(intercept = 0, slope = 1, alpha =0.3) +
  geom_point(alpha=0.3, border = NULL)+
  labs(x="Measured market price (USD/kg) ",
       y="Predicted market price (USD/kg)") +
  xlim(0, 0.75) + ylim(0, 0.75)+
  annotate("label", label= expression(paste(R^{2},'=', rsq1)),
           x=.2, y=0.7)+
  annotate("label", label=paste0("RMSE=", round(RMSE1)),
           x=.2, y=0.6)+
  theme_minimal()+
  theme(panel.grid.minor = element_blank())
print(p)
```


# Use of framework

## Scenario profitability

To test the profitability of different fertilization scenarios, we simulated four scenarios that represent current and potential investments on nitrogen fertilizer. These scenarios are:

* *ZERO*: No application of nitrogen

* *BK*: Countrywide blanket recommendation (100 N kg/ha)

* *OPyield*: Optimized to get the highest yield

* *OPnetrev*: Optimized to get the highest net revenue

These are some issues with the simulation:

We compared yields, total fertilization costs and predicted revenues from these scenarios. We also compare the average and marginal value cost ratios (AVCR and MVCR). AVCR is the average physical product (quantity of output per unit of variable input for any input level) multiplied by the value-cost ratio. An AVCR value exceeding 1 indicates profitability, while an AVCR value of 2 is often used as a shorthand criterion for economic attractiveness of an investment from the perspective of a risk-averse farmer. MVCR is the marginal physical product (the change in yield deriving from an additional kg/ha of input) and multiplied by the value-cost ratio. An MVCR value exceeding 0 indicates than a risk-neutral farmer should invest more, as marginal returns are still positive. 


$$\sf{AVCR}=\displaystyle p_Y*Y/X$$
where $p_Y$ is the price of output (USD/kg), $Y$ is the yield (kg/ha) and $X$ is the amount of fertilizer used (kg/ha).

$$\sf{MVCR}=\displaystyle \frac{p_y*(Y_F-Y_{F-1})}{p_X(X_F-X_{F-1})}$$
where $p_Y$ and $p_X$ are the prices of output and input in (USD/kg), $Y_F$ is the yield with fertilizer application (kg/ha) and $Y_{NF}$ is the yield without fertilizer application (kg/ha). $X_F$ and $X_{NF}$ are the amounts of fertilizer used for the two treatments in kg/ha. In this work, the increments of fertilizer are of 1kg/ha, therefore, the MVCR is just:
$$\sf{MVCR}=\displaystyle \frac{p_y*(Y_F-Y_{F-1})}{p_X}$$


All code for the simulation and analysis is openly available https://github.com/spalmas/ex-ante.


# Results
Table X: Summary statistics for yields, total fertilization costs and netrevenues for the different scenarios.
* Low and high net revenue comes form the 10 and 90% percentile of net revenue from the ZERO simulation
* table useful to determine visualization parameters for maps


```{r, echo=FALSE, warning=FALSE}
#Low and high netrev values will bebased on 10 and 90% percenties of the ZERO simulation
ZERO <- read.csv('../results/tables/TZA_ZERO.csv') 

#Low and high revenue from the ZERO simulation
low_netrev <- quantile(ZERO$netrev_mean, 0.10, na.rm=TRUE)  #10% percentile of the net revenue. The quantile function does not consider the Inf values.
high_netrev <- quantile(ZERO$netrev_mean, 0.90, na.rm=TRUE)  #90% percentile of the net revenue. The quantile function does not consider the Inf values. 

maize_area <- sum(ZERO$spam2010V1r1_global_A_MAIZ_A_TZA>0, na.rm=TRUE) #total area. In this case number of pixels of 5km

#start building ranges table
scenarios <- c('ZERO', 'BK', 'OPyield', 'OPnetrev')
columns <- c('Scenario',
             'yieldMin', 'yieldMean', 'yieldMax',
             'totfertcostMin', 'totfertcostMean', 'totfertcostMax',
             'netrevMin', 'netrevMean', 'netrevMax',
             paste0("Low netrev area % (<$", round(low_netrev), " USD/ha)"),
             paste0("High netrev area % (>$", round(high_netrev), " USD/ha)"))
ranges <- matrix(nrow = length(scenarios)*2 - 1, ncol = length(columns)) %>% as_tibble()
ranges[,1] <- c( 'ZERO','BK', 'OPyield', 'OPnetrev','BK_gainPerc','OPyield_mean_gainPerc','OPnetrev_mean_gainPerc') 
colnames(ranges) <- columns

#COUNTRY <- 'TZA'
r <- 1
for (s in scenarios){
  #s <- 'ZERO'  #to test
  #### \\ Reading pixel result tables  ####
  results <- paste0('../results/tables/TZA_',s,'.csv') %>% read.csv()
  col.start <- 2
  
  #calculating yield, totfertcost and netrevenus values
  for (i in c('yield_mean', 'totfertcost', 'netrev_mean')){
    #i <- 'netrev'    #to test
    #### \\ Removing infinite values ####
    results[is.infinite(results[,i]),i] <- NA
    
    #### \\ Filling table ####
    ranges[r,col.start:(col.start+2)] <- c(results[,i] %>% min(na.rm = T) %>% round, 
                                           results[,i] %>% mean(na.rm = T) %>% round,
                                           results[,i] %>% max(na.rm = T) %>% round)
    
    #### \\ Calculating percentage gain distribution from the ZERO scenario ####
    if (s != 'ZERO'){
      ranges[r + 3,col.start:(col.start+2)] <- c(results[,paste0(i,'_gainPerc')] %>% quantile(0.05, na.rm = T) %>% round, 
                                                 results[,paste0(i,'_gainPerc')] %>% quantile(0.5, na.rm = T) %>% round,
                                                 results[,paste0(i,'_gainPerc')] %>% quantile(0.95, na.rm = T) %>% round)    
    }
    #Changing column start for change in variable i
    col.start <- col.start + 3
  }
  
  #### \\  areas under and above lower and upper net revenue values ####
  ranges[r, 11] <- 100 * sum(results$netrev_mean<=low_netrev, na.rm = TRUE)/maize_area
  ranges[r, 12] <- 100 * sum(results$netrev_mean>=high_netrev, na.rm = TRUE)/maize_area
  
  #Changing row for change in scenario sJ
  r <- r + 1
}
knitr::kable(ranges)  #print
data.table::fwrite(ranges, file = '../results/tables/table1_TZA_Ranges.csv')
```

### Yield and net revenue distribution by scenario
```{r, echo=FALSE, warning=FALSE, , fig.height=3, fig.width=6}
scenarios <- c('ZERO', 'BK', 'OPyield', 'OPnetrev')
values <- tibble()

for (s in scenarios){
  #s <- 'BK'  #to test
  #### \\ Reading pixel result tables  ####
  results <- read.csv(paste0('../results/tables/TZA_',s,'.csv'))
  value <- cbind.data.frame(s, yield_mean=results$yield_mean)
  values <- rbind(values, value)
}

#Getting the percentage increase labels to add in the plot
ranges <- read.csv('../results/tables/table1_TZA_Ranges.csv') %>% 
  filter(Scenario %in% c("ZERO", "BK_gainPerc", "OPyield_mean_gainPerc", "OPnetrev_mean_gainPerc")) %>% 
  dplyr::select(Scenario, yieldMean) %>% 
  mutate(label = paste0(yieldMean, "%"))
ranges$label[1] <- NA
ranges$s <- factor(scenarios, levels=scenarios)

p <- ggplot(values, aes(x=s, y=yield_mean, fill=s)) + 
  geom_boxplot(notch=TRUE,
               outlier.size = 0.3) +
  labs(y="Yield (kg/ha)") +
  ylim(0,6000) +
  geom_label(data=ranges, aes(label=label, fill=s), color="black", x=1:4, y=5400)+
  annotate("label", label="Average Change", x=3, y=6000)+
  scale_fill_manual(values=cbbPalette) +
  theme_minimal()+
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        axis.title.x=element_blank())

####### +++++++  NET REVENUE +++++++  #######
scenarios <- c('ZERO', 'BK', 'OPyield', 'OPnetrev')
values <- tibble()

for (s in scenarios){
  #s <- 'BK'  #to test
  #### \\ Reading pixel result tables  ####
  results <- paste0('../results/tables/TZA_',s,'.csv') %>% read.csv()
  value <- cbind.data.frame(s, netrev_mean=results$netrev_mean) %>% as_tibble()
  values <- rbind(values, value)
}

#Getting the percentage increase labels to add in the plot
ranges <- read.csv('../results/tables/table1_TZA_Ranges.csv') %>% 
  filter(Scenario %in% c("ZERO", "BK_gainPerc", "OPyield_mean_gainPerc", "OPnetrev_mean_gainPerc")) %>% 
  dplyr::select(Scenario, netrevMean) %>% 
  mutate(label = paste0(netrevMean, "%"))
ranges$label[1] <- NA  #for scenario ZERO
ranges$s <- factor(scenarios, levels=scenarios)

r <- ggplot(values, aes(x=s, y=netrev_mean, fill=s)) + 
  geom_boxplot(notch =TRUE,
               outlier.size = 0.3) +
  geom_hline(yintercept = 0, color='coral', size=1) +
  labs(y="Net revenue (USD/ha)") +
  ylim(-500,3000) +
  geom_label(data=ranges, aes(label=label), color="black", x=1:4, y=2600)+
  annotate("label", label="Average Change", x=3, y=3000)+
  scale_fill_manual(values=cbbPalette) + 
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        axis.title.x=element_blank())

figureX <- ggarrange(p, r,labels=c("A", "B"),
                     ncol=2, nrow=1)
figureX


ggsave(filename = paste0('D:/OneDrive - CGIAR/Ex-Ante/Manuscript Ex-Ante/figures/Fig4_yield_netrev_boxplot.tif'),
      plot = figureX,
      width = 6, height = 3, units = 'in',
      device = 'tiff', dpi = 300)

```


### Yield area distribution
```{r, echo=FALSE, message=FALSE, warning=FALSE}
scenarios <- c('ZERO', 'BK', 'OPyield', 'OPnetrev')
values <- tibble()

for (s in scenarios){
  #s <- 'BK'  #to test
  #### \\ Reading pixel result tables  ####
  results <- paste0('../results/tables/TZA_',s,'.csv') %>% read.csv()
  value <- cbind.data.frame(s, yield_mean=results$yield_mean) %>% as_tibble()
  values <- rbind(values, value)
}

p <- ggplot(values, aes(x=yield_mean,  fill=s)) + 
  geom_density(alpha=0.5, size=0.3) +
  scale_fill_manual(values=cbbPalette) +
  xlab("Yield (kg/ha)") +
  ylab(bquote('Area (km'^2*')')) +
  theme_minimal()+
  theme(legend.position="bottom",
        legend.title = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
print(p)
```

### Net revenue area distribution
```{r, echo=FALSE, message=FALSE, warning=FALSE}
scenarios <- c('ZERO', 'BK', 'OPyield', 'OPnetrev')
values <- tibble()

for (s in scenarios){
  #s <- 'BK'  #to test
  #### \\ Reading pixel result tables  ####
  results <- paste0('../results/tables/TZA_',s,'.csv') %>% read.csv()
  value <- cbind.data.frame(s, netrev_mean=results$netrev_mean) %>% as_tibble()
  values <- rbind(values, value)
}

p <- ggplot(values, aes(x=netrev_mean , fill=s)) + 
  geom_density(alpha=0.5, size=0.3) +
  scale_fill_manual(values=cbbPalette) +
  xlab("Net revenue (USD/ha)") +
  ylab(bquote('Area (km'^2*')')) +
  theme_minimal()+
  theme(legend.position="bottom",
        legend.title = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
print(p)
```



### Table 2: Percentage of areas with high profitability by scenario by district
This estimate is only areas covered with maize

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#Low and high netrev values will bebased on 10 and 90% percenties of the ZERO simulation
ZERO <- read.csv('../results/tables/TZA_ZERO.csv') 
high_netrev <- quantile(ZERO$netrev_mean, 0.90, na.rm=TRUE) %>% round()  #90% percentile of the net revenue 

scenarios <- c('ZERO', 'BK', 'OPyield', 'OPnetrev')
cols <- length(scenarios)  + 2

#### \\ Getting total area of maize per adm_1 and AEZ
FIDS_gadm36_TZA_1 <- unique(ZERO$gadm36_TZA_1) %>% sort()  
n_gadm36_TZA_1 <- length(FIDS_gadm36_TZA_1)  #how many regions
pixel_area <- 5*5  #pixel size (km2) 
area_gadm36_TZA_1 <- table(ZERO$gadm36_TZA_1) * pixel_area  #Area of each region

#table to get the district names
gadm36_TZA_1 <- read_csv("../data/admin/gadm36_TZA_1.csv") 

#### \\ Creating matrix ####
results <- matrix(nrow = n_gadm36_TZA_1, ncol = cols)
colnames(results) = c('District',
                      'Maize Area (km2)',
                      paste0('ZERO (>$', high_netrev ,')'),
                      paste0('BK (>$', high_netrev ,')'),
                      paste0('OPyield (>$', high_netrev ,')'),
                      paste0('OPnetrev ($', high_netrev ,')'))
results <- as_tibble(results)

#### \\ adding level_1 name ####
results[1:n_gadm36_TZA_1, 1] <- gadm36_TZA_1$NAME_1[match(FIDS_gadm36_TZA_1, gadm36_TZA_1$FID)] 

#### \\ adding area####
results[1:n_gadm36_TZA_1, 2] <- area_gadm36_TZA_1

#### \\ scenario results ####
co <- 3  #starting column to store results
for (s in scenarios){
  #s <- 'ZERO'  #to test
  pixel_results <- paste0('../results/tables/TZA_', s,'.csv') %>% read.csv()
  
  #changing gadm36_TZA_1 to factor characters
  pixel_results$gadm36_TZA_1 <- pixel_results$gadm36_TZA_1 %>% factor()
  
  #Number of pixels with high profitability
  pixel_high_netrev <- pixel_results$netrev_mean >= high_netrev
  
  #areas of each gadm36_TZA_1 and aez with high profitability
  results[1:n_gadm36_TZA_1,co] <- pixel_area * pixel_results$gadm36_TZA_1[pixel_high_netrev] %>% table()
  
  co <- co + 1
}

#### \\ Adding a total row at the end ####
results <- rbind(results, c(NA, colSums(results[,2:cols])))
results[nrow(results),1] <- "TOTAL"

#### \\ Convert to percentages ####
results[,3:cols] <- round(100 * results[,3:cols]/results[,2,drop = T])

#### \\ Printing table ####
knitr::kable(results)

#### \\ finishing touches ####
write_csv(results, path = paste0('../results/tables/table2_TZA.csv'))

```

### Table X: Summary table of aggregate gains in net revenue
```{r, message=FALSE, echo=FALSE, warning=FALSE}
#regions of tanzania
gadm36_TZA_1 <- vect("F:/Work/GADM/gadm36_levels_shp/gadm36_TZA_shp/gadm36_TZA_1.shp")

#starts of results table
results <- data.frame(ID = gadm36_TZA_1$ID, Region = gadm36_TZA_1$NAME_1)

#Rural population
pop2020 <- rast("F:/Work/WorldPop/Africa/AFR_PPP_2020_adj_v2.tif")
results$pop2020 <- as.vector(terra::extract(pop2020, gadm36_TZA_1, fun = sum, na.rm=TRUE, drop=TRUE))
urban_perc <- readxl::read_xlsx("F:/Work/Tanzania/rural-urban_population/muzzini-WB.xlsx", sheet = "data") %>% select(Region, Statistical) %>% rename(urban_perc = Statistical)
results <- left_join(results, urban_perc, by = "Region")
results$rural_pop2020 <- results$pop2020 * (100-results$urban_perc)/100

#SPAM to get the hectares under maize production
SPAM2010 <- rast("F:/Work/SPAM/spam2010v1r1/spam2010v1r1_global_phys_area.geotiff/spam2010V1r1_global_A_MAIZ_A.tif")
results$Maize_ha <- as.vector(terra::extract(SPAM2010, gadm36_TZA_1, fun = sum, na.rm=TRUE, drop=TRUE))

#Expected total returns USD/ha*ha of maize if move from ZERO to OPnetrev and from BK to OPnetrev
OPnetrev <- read_csv("../results/tables/TZA_OPnetrev.csv")

OPnetrev_ZERO <- read_csv("../results/tables/TZA_ZERO.csv") %>%
  mutate(OPnetrev_ZERO_netrev_gain = OPnetrev$netrev_mean - netrev_mean) %>% 
  group_by(gadm36_TZA_1) %>% #these are from the ID values in the rasterize in preprocess_rasters.sh
  summarise(netrev_gain = mean(OPnetrev_ZERO_netrev_gain)) %>% 
  rename(ID = gadm36_TZA_1)
results$OPnetrev_ZERO_netrev_gain <- left_join(results, OPnetrev_ZERO, by="ID")$netrev_gain

OPnetrev_BK <- read_csv("../results/tables/TZA_BK.csv") %>% 
  mutate(OPnetrev_BK_netrev_gain = OPnetrev$netrev_mean - netrev_mean) %>% 
  group_by(gadm36_TZA_1) %>% #these are from the ID values in the rasterize in preprocess_rasters.sh
  summarise(netrev_gain = mean(OPnetrev_BK_netrev_gain)) %>% 
  rename(ID = gadm36_TZA_1)
results$OPnetrev_BK_netrev_gain <- left_join(results, OPnetrev_BK, by="ID")$netrev_gain


#selecting and rounding result for printing
results <- results %>% select(Region, rural_pop2020, Maize_ha, OPnetrev_ZERO_netrev_gain, OPnetrev_BK_netrev_gain) %>% 
  filter(!is.na(OPnetrev_BK_netrev_gain)) %>% 
  mutate(rural_pop2020 = round(rural_pop2020/1000000,digits = 1),  #million peopel
         Maize_ha = round(Maize_ha/100,digits=1),  #to convert ha to km2
         OPnetrev_ZERO_netrev_gain = round(OPnetrev_ZERO_netrev_gain, digits=1),
         OPnetrev_BK_netrev_gain = round(OPnetrev_BK_netrev_gain, digits=1))


####  Printing table ####
knitr::kable(results)

#### finishing touches ####
write_csv(results, path = "D:/OneDrive - CGIAR/Ex-Ante/Manuscript Ex-Ante/tables/tableX_aggregateGains.csv")

```

# Percentage of pixels within ranges
```{r}
OPnetrev_netrev_cv <-  rast("../results/tif/TZA_OPnetrev_netrev_cv.tif")


print(">10")
sum(values(OPnetrev_netrev_cv)>5, na.rm=TRUE)/sum(values(OPnetrev_netrev_cv)>0, na.rm=TRUE)

hist(values(OPnetrev_netrev_cv))

```


# Percentage of pixels within ranges: Nitrogen amount optimized
```{r}
r <-  rast("../results/tif/TZA_OPnetrev_N_kgha.tif")
sum(values(r)>50 & values(r)<100, na.rm=TRUE)/sum(values(r)>0, na.rm=TRUE)
hist(values(r))

```

