---
title: "Random Forest Prediction of yield response"
#author: "Sebastian Palmas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

# Packages
```{r, message=FALSE}
library(tidyverse)
```

# Data
This is the TZAPS yield measurements from 2016 and 2017 added with many layers of information from ISRIC, markus stacks, access from Jordan, CHIRPS season data (from 2016 and 2017), CGIAR-srtm data.
This file was prepared in the *TZAPS_hhid_coords_extract.R* code.
```{r, message=FALSE}
TZAPS <- read_csv("F:/Work/TAMASA/APS/TZAPS_hhid_coords_stacks.csv") %>% 
#TZAPS <- read_csv("/media/sp/SPHD/Work/TAMASA/APS/TZAPS_hhid_coords_stacks.csv") %>%    #if working in linux
  filter(!is.na(yld)) %>%    #remove plots without yield data
  filter(yld > 100) %>%     #outliers?
  filter(TAMSAT_seas_rainfall > 300) %>%   # Deleting places where seasonal rainfall is too low
  dplyr:: select(-K_kgha) %>%   #Probably not measured correctly
  dplyr::select(-hhid, -district) %>%   #household ID and district should not be in the model
  dplyr::select(-latitude, -longitude) %>%   #repeated columns in the Markus stack
  dplyr::select(-sloped) %>% #better to use our own elevation and slope raster
  mutate(highlands = as.factor(highlands)) %>% #RF does not like character variables
  dplyr::select(-highlands) %>% #not significant, can be correlated with coordinates
  #select(-CEC, -PH, -SND, -SOC) %>% #repeated SoilGrids layers, I will use I have in files.isric
  #select(-WPOP) %>% #we already have improved population estimate from Worldpop
  #dplyr::select(-acc100k, -acc50k, -accdar, -accprd) %>%    #access variables in the yield model?
  dplyr::select(-BIO15, -TAMSAT_annual_rainfall) #%>%   #Delete rainfall columns
  #rename(annual_rainfall = BIO12)  #the one that will be used for rainfall
  #select(-TAMSAT_seas_rainfall) %>%   #incomplete data
  #select(-CHIRPS_seas_rainfall) %>%   #to avoid autocorrelation

  

#We filter all the observations without yield. 
head(TZAPS)
```
# CHIRPS vs TAMSAT
TAMSAT has considerably lower rainfall estimates:

```{r, echo=FALSE}
plot(TZAPS$CHIRPS_seas_rainfall , TZAPS$TAMSAT_seas_rainfall)
abline(a=0,b=1)
```


# lm Yield prediction

```{r}
lm1 <- lm(yld~., TZAPS)
summary(lm1)
```

## What about just some of the variables
With CHIRPS data
```{r}
lm2 <- lm(yld ~ N_kgha  + year + intercrop + rotat1 + manure + cropres + weedings + herbicide_bin + pesticide_bin + impseed + disease + striga + logha + headage + femhead + hhsize + headeduc + srtm_TZA + srtm_slope_TZA + CHIRPS_seas_rainfall, data = TZAPS)
summary(lm2)
```
With TAMSAT data
```{r}
lm3 <- lm(yld ~ N_kgha  + year + intercrop + rotat1 + manure + cropres + weedings + herbicide_bin + pesticide_bin + impseed + disease + striga + logha + headage + femhead + hhsize + headeduc + srtm_TZA + srtm_slope_TZA + TAMSAT_seas_rainfall, data = TZAPS)
summary(lm3)
```




# Random Forest Yield prediction
We need to check if there are any predictors with NA values
```{r}
for(column in seq_along(TZAPS)){
  if(any(is.na(TZAPS[column]))){
    print(paste0("Column: ", colnames(TZAPS)[column], " has at least one NA value"))
  }
}
```
For those columns, let's do something about it
```{r}
#We can't do anything if there is no fertilizer data. It is one row I think. Let's delete it. (Should we use zeroes for those numbers?)
TZAPS <- TZAPS %>% filter(!is.na(N_kgha))
#column N_kgha,P_kgha,K_kgha and logha  are now without problems

#for column headage and headagesq let's fill 22 with the mean age from the rest of the table and therefore fill headagesq with their squares
TZAPS$headage[is.na(TZAPS$headage)] <- mean(TZAPS$headage, na.rm=TRUE)
TZAPS$headagesq <- TZAPS$headage^2

# Similarly, let's fill femhead and headeduc with the mean from the rest
TZAPS$femhead[is.na(TZAPS$femhead)] <- mean(TZAPS$femhead, na.rm=TRUE)
TZAPS$headeduc[is.na(TZAPS$headeduc)] <- mean(TZAPS$headeduc, na.rm=TRUE)

```


Let's load the RF library
```{r, message=FALSE}
library(randomForest)
```
## Model 1
### Tune The Forest
By "tune the forest", we mean the process of determining the optimal number of variables to consider at each split in a decision-tree. Too many prediction variables and the algorithm will over-fit; too few prediction variables and the algorithm will under-fit. so first, we use `tuneRF` function to get the possible optimal numbers of prediction variables. The `tuneRF` function takes two arguments: the prediction variables and the response variable.
```{r}
trf <- tuneRF(x=TZAPS[,2:ncol(TZAPS)], # Prediction variables
              y=TZAPS$yld) # Response variable
```

`tuneRF` returns the several numbers of variables randomly sampled as candidates at each split (mtry). To build the model, we pick the number with the lowest [Out-of-Bag (OOB) error.](https://en.wikipedia.org/wiki/Out-of-bag_error) prediction error.
```{r}
(mintree <- trf[which.min(trf[,2]),1])
```


### Fit The Model
We create a model with the `randomForest` function which takes as arguments: the response variable the prediction variables and the optimal number of variables to consider at each split (estimated above). We also get the function to rank the prediction variables based on how much influence they have in the decision-trees' results.
```{r}
yield.rf <- randomForest(yld~., TZAPS,
                         mtry=mintree, # Number of variables in subset at each split
                         importance = TRUE) # Assess importance of predictors.
yield.rf
```

We can have a look at the model in detail by plotting it to see a plot of the number of trees against OOB error: the error rate as the number of trees increase. 
```{r}
plot(yield.rf)
```

RMSE of the optimal random forest
```{r}
(oob <- sqrt(yield.rf$mse[which.min(yield.rf$mse)]))
```
We can have a look at each variable's influence by plotting their importance based on different indices given by the [importance function](https://www.rdocumentation.org/packages/randomForest/versions/4.6-14/topics/importance).
```{r,fig.width=10, fig.height=6}
varImpPlot(yield.rf)
```

### Prediction Evaluation
To assess the predictions, we will compare the observed maize yields (the training data) with the predicted maize yields (predicted using the training data). Considering that the training data, is in point form as opposed to raster, we will make a non-spatial prediction using the predict function in the "stats" package. We plot the observed and predicted values to see the trend. 
```{r, echo=FALSE}
pred <- stats::predict(yield.rf)

rsq <- function (obs, pred) cor(obs, pred, use = 'complete.obs') ^ 2
RMSE <- function(obs, pred){sqrt(mean((pred - obs)^2, na.rm = TRUE))}
plot_fitness <- function(obs, pred, name){
  r2 <- rsq(obs, pred) %>% round(digits = 3)
  rmse <- RMSE(obs, pred) %>% round(digits = 3)
  
  plot(x = obs, pred,
       main = paste0(name, ". Quick r2 = ", r2, ". RMSE = ", rmse))
  abline(a=0, b=1)
}

plot_fitness(obs = TZAPS$yld,
             pred = pred,
             "Fitness")
```
### Partial dependence plots
Partial dependence plots of selected variables
```{r}
library(caret)
#plotting only the 10 most important variables
impvar <- rownames(varImp(yield.rf))[order(varImp(yield.rf), decreasing = TRUE)]
impvar <- impvar[1:10]
op <- par(mfrow=c(2, 5))
for (i in seq_along(impvar)) {
    partialPlot(yield.rf, as.data.frame(TZAPS), impvar[i], xlab=impvar[i],
                main="Partial Dependence")
}
par(op)
```
Plotting dependence using the `pdp` package.
```{r}
library(pdp)
pdp::partial(yield.rf, pred.var="TAMSAT_seas_rainfall", plot=TRUE)
```

We can save the model to a file and then load it with `readRDS` to use it with another set of predictors.
```{r}
saveRDS(yield.rf, "../results/models/yield.rf.rds")
```


## Random Forest model 2
In this model, I'll remove many of the predictors that do not have much information about yield. This is to reduce the size of the file that will be used to predict.

### Select the columns in the model
```{r}
TZAPS_subset <- TZAPS %>% dplyr::select(yld, N_kgha, intercrop, rotat1, manure,
                                        cropres, weedings, impseed, fallow3yr,
                                        struct, terraced, logha, headage, hhsize,
                                        headeduc, year, srtm_TZA, srtm_slope_TZA,
                                        af_ORCDRC_T__M_sd1_5000m_TZA,#Longitude, Latitude,
                                        af_PHIHOX_T__M_sd1_5000m_TZA,
                                        TAMSAT_seas_rainfall) %>% 
  rename(seas_rainfall = TAMSAT_seas_rainfall)
```


### Tune The Forest
```{r}
set.seed(1234)
trf <- tuneRF(x=TZAPS_subset[,2:ncol(TZAPS_subset)], # Prediction variables
              y=TZAPS_subset$yld) # Response variable
```

`tuneRF` returns the several numbers of variables randomly sampled as candidates at each split (mtry). To build the model, we pick the number with the lowest [Out-of-Bag (OOB) error.](https://en.wikipedia.org/wiki/Out-of-bag_error) prediction error.
```{r}
(mintree <- trf[which.min(trf[,2]),1])
```


### Fit The Model
We create a model with the `randomForest` function which takes as arguments: the response variable the prediction variables and the optimal number of variables to consider at each split (estimated above). We also get the function to rank the prediction variables based on how much influence they have in the decision-trees' results.
```{r}
yield.rf2 <- randomForest(yld~., TZAPS_subset,
                          mtry=mintree, # Number of variables in subset at each split
                          importance = TRUE) # Assess importance of predictors.
yield.rf2
```

We can have a look at the model in detail by plotting it to see a plot of the number of trees against OOB error: the error rate as the number of trees increase. 
```{r}
plot(yield.rf2)
```

RMSE of the optimal random forest
```{r}
(oob <- sqrt(yield.rf2$mse[which.min(yield.rf2$mse)]))
```
We can have a look at each variable's influence by plotting their importance based on different indices given by the [importance function](https://www.rdocumentation.org/packages/randomForest/versions/4.6-14/topics/importance).
```{r,fig.width=10, fig.height=6}
varImpPlot(yield.rf2)
```
### Partial dependence plots

Partial dependence plots of the 10 most important variables.

```{r}
library(caret)
impvar <- rownames(varImp(yield.rf2))[order(varImp(yield.rf2), decreasing = TRUE)]
impvar <- impvar[1:10]
op <- par(mfrow=c(2, 5))
for (i in seq_along(impvar)) {
    partialPlot(yield.rf2, as.data.frame(TZAPS_subset), impvar[i], xlab=impvar[i],
                main="Partial Dependence")
}
par(op)
```


In here we also build the partial dependence plots that will be used in the paper.
```{r}
library(ggpubr)   # for theme and ggarrange package
theme_set(theme_pubr())


seas_rainfall_partial <- partialPlot(x=yield.rf2, pred.data = as.data.frame(TZAPS_subset),
                                     x.var = "seas_rainfall", plot=FALSE) %>% data.frame()
N_kgha_partial <- partialPlot(x=yield.rf2, pred.data = as.data.frame(TZAPS_subset),
                              x.var = "N_kgha", plot=FALSE) %>% data.frame()
ORCDRC_partial <- partialPlot(x=yield.rf2, pred.data = as.data.frame(TZAPS_subset),
                              x.var = "af_ORCDRC_T__M_sd1_5000m_TZA", plot=FALSE) %>% data.frame()

seas_rainfall_plot <- ggplot(seas_rainfall_partial, aes(x,y)) +
  geom_line() +
  labs(x = "Seasonal rainfall (mm)", y = "Partial yield (kg/ha)")
N_kgha_plot <- ggplot(N_kgha_partial, aes(x,y)) +
  geom_line() +
  labs(x = "Nitrogen (kg/ha)", y = "Partial yield (kg/ha)")
ORCDRC_plot <- ggplot(ORCDRC_partial, aes(x,y)) +
  geom_line() +
  labs(x = "Soil Organic carbon (g/kg)", y = "Partial yield (kg/ha)")
```


### Prediction Evaluation
To assess the predictions, we will compare the observed maize yields (the training data) with the predicted maize yields (predicted using the training data). Considering that the training data, is in point form as opposed to raster, we will make a non-spatial prediction using the predict function in the "stats" package. We plot the observed and predicted values to see the trend. 
```{r, echo=FALSE, fig.length = 4, fig.width = 4}
TZAPS_subset$pred <- stats::predict(yield.rf2)

rsq <- function (obs, pred) cor(obs, pred, use = 'complete.obs') ^ 2
RMSE <- function(obs, pred){sqrt(mean((pred - obs)^2, na.rm = TRUE))}

fr2_rsq <- rsq(TZAPS_subset$yld, TZAPS_subset$pred) %>% round(digits = 2)
fr2_rmse <- RMSE(TZAPS_subset$yld, TZAPS_subset$pred) %>% round(digits = 0)

yield_fit_plot <- ggplot(data = TZAPS_subset, aes(x = yld, y = pred)) +
  geom_point(alpha=0.2) +
  geom_abline(slope = 1, alpha=0.3) +
  annotate('text', x = 2000, y = 7500, label = paste0("R^{2}==", fr2_rsq), parse = TRUE, size=3)  +
  annotate('text', x = 2000, y = 6500, label = paste0("RMSE==", fr2_rmse), parse = TRUE, size=3)  +
  labs(x = "Observed yield (kg/ha)", y = "Predicted yield (kg/ha)") +
  xlim(0, 9000) + ylim(0, 9000)
```

### Building the figure with the 4 plots
```{r}
figure2 <- ggarrange(seas_rainfall_plot, N_kgha_plot, ORCDRC_plot, yield_fit_plot,
                     labels=c("A", "B", "C", "D"),
                     ncol=2, nrow=2)
figure2
ggsave(filename = "D:/OneDrive - CGIAR/Ex-Ante/Manuscript Ex-Ante/figures/Figure3_yieldfit.tif",
       plot = figure2,
       width = 15, height = 15, units = "cm", device = "tiff")
```


We can save the model to a file and then load it with `readRDS` to use it with another set of predictors.
```{r}
saveRDS(yield.rf2, "../results/models/yield.rf2.rds")
save(yield.rf2, file="../results/models/yield.rf2.rda")
```

# Mean values of household variables
Because we don't have spatial surfaces of the household variables, we will use the mean values of these variables in the TZAPS as constants in the spatial predictions. This code below finds the mean values of the household characteristics found in the TZAPS.

The exported table will be used to fill the values in the simulations.
```{r}
hh_summary_stats <- TZAPS_subset %>% 
  dplyr::select(intercrop, rotat1, manure, cropres, weedings, impseed, fallow3yr, struct, terraced, logha, headage, hhsize, headeduc, year) %>%
  sapply(FUN = function(cl) c(mean=mean(cl,na.rm=TRUE), 
                              sd=sd(cl,na.rm=TRUE))) %>% 
  t() %>% data.frame() %>% rownames_to_column("variable")

hh_summary_stats
write.table(hh_summary_stats,file="../data/hh_summary_stats.csv", sep=",", row.names = FALSE)
print(paste0("Number of observations: ", nrow(TZAPS_subset)))

```